\section{Model}

We first specify a formal model for concurrency below. We provide a high-level
overview and then show how it can be applied to various productions within the P4
grammar. The compiler section of this draft discusses compilation strategies.

An instance of a parser or control block type (\S9.2.8 of the spec) is called a
parser or control block instance.  All computations within P4 happen inside
these instances as specified by the statements (\S12) within them.  We say a
parser or control block is ``invoked'' when some external event triggers the
execution of a parser instance or a control block instance.  This external
event could be the arrival of an unparsed packet (\texttt{packet\_in}) for a
parser instance or parsed headers (\texttt{in} or \texttt{inout} function
arguments) for a control block instance.

When an event invokes a parser or control block instance, we can conceptually
think of the instance launching a new thread to handle that event in the
background.  The instance is now free to handle the next event. The concurrency
model doesn't specify the implementation. For instance, the number of threads
in a thread pool is implementation-defined.  In fact, the implementation
doesn't even need to use a thread pool. It can instead use a processing
pipeline, where a small portion of the work for each event is handled by one
thread, which then passes off the event to the next thread (e.g., match-action
pipelines).

Each of the launched threads runs to completion, executing all statements
within the control block or parser instance one after another, sequentially.
Once the thread completes processing an event within an instance, it triggers
an event within another parser or control block instance, as specified by the
target architecture.

Any interleaving of threads is permitted as long as statements within a thread
aren't reordered. The atomic (or instantaneous) units of execution within each
thread's statements are defined by providing an \texttt{@atomic} annotation
around specific statements or expressions in the program.\footnote{An
expression can be treated as a statement by assigning the result of the
expression to a temporary variable.} This annotation denotes that the
computations within those statements or expressions appear to execute
atomically/instantaneously.

In practice, the P4 compiler annotates many expressions and statements atomic
by default, to reduce the burden on the P4 programmer. We dicuss these below.

\subsection{atomic by default}
Statements and expressions that are atomic by default are all statements and
expressions that do not access an extern instance, i.e., statements or
expressions that read or write the following types of data:
\begin{enumerate}
\item intrinsic metadata (\S4)
\item metadata (\S4)
\item packet headers (\S4)
\item local variables declared within a control block's apply block, a parser's
state, or an action's body (\S11.2)\footnote{Right now, P4 doesn't permit
extern instances as local variables. The FAQ section of this draft expands on
this.}.

\end{enumerate}
We also include statements that recursively only include statements that read
or write the above types of data. An example is a statement that calls an action,
which respects the conditions above.
%Note that this is better than saying actions are atomic by default, because
%that might be too much to begin with.

The rationale for not including extern instances in the atomic-by-default list above
 is that extern instances
encapsulate data-plane state, unlike the scenarios above that don't include any
state visible to all threads. State modification through extern method calls
cannot be declared atomic by default because depending on the extern method,
some intermediate state may be visible to another thread, which contradicts the
requirement that modifications be instantaneous. We recommend that method calls
on extern instances be implemented as atomic by target architecture
implementers, e.g., serializing all calls to a counter or register to provide
the illusion of atomicity. An atomic method call can be specified by using the
\texttt{@atomic} notation within an extern declaration (the next section provides
an example).

Certain combination of statements are also annotated atomic by default. As an
example, consider the statements:
\begin{verbatim}
x = x & 4w0;
\end{verbatim}

and
\begin{verbatim}
x = 4w0 & x;
\end{verbatim},
where x is a local variable. Then the block statement:

\begin{verbatim}
{
  x = x & 4w0;
  x = 4w0 & x;
}
\end{verbatim}
is atomic by default because it calls no extern instance.

Certain categories of statements are explicitly not atomic by default. In
particular, while it is recommended that individual method calls on extern
instances be atomic, a pair of calls to two different extern instances is not
atomic by default. An example is a pair of counters, say, \texttt{c1} and
\texttt{c2}.  While the increment method call for each counter appears to
execute atomically, other statements can be interleaved between the increments
to \texttt{c1} and \texttt{c2}, unless the programmer encloses both method
calls within one atomic annotation.

The above discussion should give a sense for how atomic is used in practice. To
reduce programmer burden, we recommend that P4 compilers implement a inference
procedure for conservatively inferring whether any given production within a
grammar is atomic by default. The procedure could be modeled based on some of
the rules given above.
%TODO: We would ideally provide a full inference procedure in the appendix.
